---
title: "SCIR 297_Food Access_Final Project"
author: "Catherine Z. Matenje"
format: html
editor: visual
---

## Introduction

This project uses data from the USDA Food Access Research Atlas (2019) to examine food access across U.S. census tracts. The dataset provides information on food access, socioeconomic conditions, and transportation characteristics for U.S. census tracts. It includes measures of low food access, poverty, income, vehicle availability, and SNAP participation, which can be used to examine factors associated with food access across communities. For this project, variables related to food access and community vulnerability were selected for analysis.

My original research interest focused on disaster food insecurity, specifically how disasters affect household food security before, during, and after disaster events. Because the disaster-specific dataset was not available in time for this project, this analysis looks at food access instead as a related measure of community vulnerability to disasters. Communities with existing barriers to food access may face greater challenges during emergencies or disasters when transportation, food supply chains, and local infrastructure are disrupted.

For this exploratory data analysis, I am considering two possible research questions.

Research Question 1: What socioeconomic and transportation factors are associated with low food access in U.S. census tracts?

| Variable | Description | Outcome/predictor |
|------------------------|------------------------|------------------------|
| LILATracts_1And10 | Indicates whether a census tract is classified as both low-income and low-access (food desert indicator). | Outcome Variable |
| PovertyRate | Percentage of the population living below the federal poverty line within the census tract. | Predictor Variable |
| MedianFamilyIncome | Median family income within the census tract, representing the midpoint of the income distribution. | Predictor Variable |
| Urban | Indicates whether the census tract is classified as urban (1) or rural (0). | Predictor Variable |
| HUNVFlag | Indicates low vehicle access, identifying tracts where many households do not have access to a vehicle and may face transportation barriers to reaching supermarkets. | Predictor Variable |
| lasnap1share | Share of housing units receiving SNAP benefits that are located more than one mile from a supermarket. | Predictor Variable |

Research Question 2: Which community characteristics may increase vulnerability to food access disruptions during emergencies and disasters?

| Variable | Description | Outcome/Predictor |
|------------------------|------------------------|------------------------|
| lapop1share | Proportion of the population living more than one mile from a supermarket. This serves as a continuous measure of low food access. | Outcome Variable |
| PovertyRate | Percentage of the population living below the federal poverty line within the census tract. | Predictor Variable |
| MedianFamilyIncome | Median family income within the census tract, representing the midpoint of the income distribution. | Predictor Variable |
| Urban | Indicates whether the census tract is classified as urban (1) or rural (0). | Predictor Variable |
| HUNVFlag | Indicates low vehicle access, identifying tracts where many households do not have access to a vehicle and may face transportation barriers to reaching supermarkets. | Predictor Variable |

## Load Libraries

```{r}

# Load required libraries for data wrangling, visualization, and exploratory analysis

library(tidyverse)
library(RColorBrewer)
library(corrplot)

setwd("C:/Users/cathe/OneDrive/Desktop/Montgomery College Transition/2025-2026 MONTGOMERY COLLEGE TRANSITION/MC COURSES 25-26/Summer 2026/SCIR 297/Final Project")

```

## Load Data Set

```{r}
food_data <- read_csv("FoodAccessResearchAtlasData2019_clean.csv")

head(food_data)
dim(food_data)
names(food_data)
```

## Data Cleaning

```{r}
## Data Cleaning

# Examine structure of dataset

head(food_data)

# View dimensions
dim(food_data)

# View summary statistics
summary(food_data)
```
## Creating cleaned data set and renaming variables

```{r}
# Rename variables to shorter, more descriptive names

clean_data <- food_data |>
  rename(
    urban = Urban,
    poverty_rate = PovertyRate,
    median_income = MedianFamilyIncome,
    low_vehicle_access = HUNVFlag,
    food_desert = LILATracts_1And10,
    low_access_share = lapop1share,
    snap_low_access_share = lasnap1share,
    snap_low_access_count = lasnap1,
    low_vehicle_low_access_share = lahunv1share,
    low_income_low_access_share = lalowi1share,
    child_low_access_share = lakids1share,
    senior_low_access_share = laseniors1share,
    low_income_tract = LowIncomeTracts
  )
clean_data <- clean_data |>
  mutate(

    # Convert categorical variables
    urban = as.factor(urban),
    low_vehicle_access = as.factor(low_vehicle_access),
    food_desert = as.factor(food_desert),
    low_income_tract = as.factor(low_income_tract),

    # Convert continuous variables
    median_income = as.numeric(median_income),
    low_access_share = as.numeric(low_access_share),
    snap_low_access_share = as.numeric(snap_low_access_share),
    snap_low_access_count = as.numeric(snap_low_access_count),
    low_vehicle_low_access_share = as.numeric(low_vehicle_low_access_share),
    low_income_low_access_share = as.numeric(low_income_low_access_share),
    child_low_access_share = as.numeric(child_low_access_share),
    senior_low_access_share = as.numeric(senior_low_access_share)
  )

glimpse(clean_data)
summary(clean_data)


```

# Exploratory Data Analysis

## Research Question 1: What socioeconomic and transportation factors are associated with low food access in U.S. census tracts?

For Research Question 1, the outcome variable is food_desert, which indicates whether a census tract is classified as both low-income and low-access. The predictor variables include poverty rate, median family income, urban classification, low vehicle access, and SNAP low-access share.

#### Bar Chart: Food Desert Classification

```{r}
ggplot(clean_data, aes(x = food_desert, fill = food_desert)) +
  geom_bar() +
  scale_x_discrete(
    labels = c(
      "0" = "Not Food Desert",
      "1" = "Food Desert"
    )
  ) +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    title = "Distribution of Food Desert Classification",
    x = "Food Desert Status",
    y = "Count",
    caption = "Food desert classification indicates low-income areas with limited access to supermarkets."
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
#### Histogram: Poverty Rate

```{r}
# Create a histogram to examine poverty rates

ggplot(clean_data, aes(x = poverty_rate)) +
  geom_histogram(
    fill = brewer.pal(5, "Pastel1")[3],
    color = "black",
    bins = 30
  ) +
  labs(
    title = "Distribution of Poverty Rate Across Census Tracts",
    x = "Poverty Rate (%)",
    y = "Count",
    caption = "Poverty rate represents the percentage of individuals living below the federal poverty line."
  ) +
  theme_minimal()
```
Histogram: Median Family Income

```{r}
# Create a histogram to examine median family income

ggplot(clean_data, aes(x = median_income)) +
  geom_histogram(
    fill = brewer.pal(5, "Pastel1")[4],
    color = "black",
    bins = 30
  ) +
  labs(
    title = "Distribution of Median Family Income",
    x = "Median Income (USD)",
    y = "Count",
    caption = "Median income represents the midpoint of household income distribution within each census tract."
  ) +
  theme_minimal()
```
#### Bar Chart: Urban vs. Rural Census Tracts

```{r}
# Create a bar chart to examine urban and rural census tracts

ggplot(clean_data, aes(x = urban, fill = urban)) +
  geom_bar() +
  scale_x_discrete(
    labels = c(
      "0" = "Rural",
      "1" = "Urban"
    )
  ) +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    title = "Distribution of Urban and Rural Census Tracts",
    x = "Urban Classification",
    y = "Count",
    caption = "Urban classification indicates whether a census tract is urban or rural."
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
#### Bar Chart: Low Vehicle Access

```{r}
# Create a bar chart to examine low vehicle access

ggplot(clean_data, aes(x = low_vehicle_access, fill = low_vehicle_access)) +
  geom_bar() +
  scale_x_discrete(
    labels = c(
      "0" = "No Low Vehicle Access",
      "1" = "Low Vehicle Access"
    )
  ) +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    title = "Distribution of Low Vehicle Access",
    x = "Vehicle Access Status",
    y = "Count",
    caption = "Low vehicle access indicates transportation barriers that may affect access to supermarkets."
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
#### Histogram of SNAP Low Access Share

```{r}
# Create a histogram to examine SNAP households with low food access

ggplot(clean_data, aes(x = snap_low_access_share)) +
  geom_histogram(
    fill = brewer.pal(5, "Pastel1")[2],
    color = "black",
    bins = 30
  ) +
  labs(
    title = "Distribution of SNAP Households with Low Food Access",
    x = "SNAP Low Access Share (%)",
    y = "Count",
    caption = "Represents the share of SNAP households living more than one mile from a supermarket."
  ) +
  theme_minimal()
```

#### Boxplot: Poverty Rate by Food Desert Status

```{r}
# Compare poverty rates by food desert status

ggplot(clean_data,
       aes(x = food_desert,
           y = poverty_rate,
           fill = food_desert)) +
  geom_boxplot() +
  scale_x_discrete(
    labels = c(
      "0" = "Not Food Desert",
      "1" = "Food Desert"
    )
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Poverty Rate by Food Desert Classification",
    x = "Food Desert Status",
    y = "Poverty Rate (%)",
    caption = "This plot compares poverty rates between food desert and non-food desert census tracts."
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

#### Box Plot: Median Income by Food Desert Status

```{r}
# Compare median income by food desert status

ggplot(clean_data,
       aes(x = food_desert,
           y = median_income,
           fill = food_desert)) +
  geom_boxplot() +
  scale_x_discrete(
    labels = c(
      "0" = "Not Food Desert",
      "1" = "Food Desert"
    )
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Median Income by Food Desert Classification",
    x = "Food Desert Status",
    y = "Median Income (USD)",
    caption = "This plot compares median income between food desert and non-food desert census tracts."
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
#### Boxplot: SNAP Low Access Share by Food Desert

```{r}
# Compare SNAP low access share by food desert status

ggplot(clean_data,
       aes(x = food_desert,
           y = snap_low_access_share,
           fill = food_desert)) +
  geom_boxplot() +
  scale_x_discrete(
    labels = c(
      "0" = "Not Food Desert",
      "1" = "Food Desert"
    )
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "SNAP Low Access Share by Food Desert Classification",
    x = "Food Desert Status",
    y = "SNAP Low Access Share (%)",
    caption = "This plot compares SNAP low-access share between food desert and non-food desert census tracts."
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
## Research Question 2:Which community characteristics may increase vulnerability to food access disruptions during emergencies and disasters?

For Research Question 2, the outcome variable is low_access_share, which measures the percentage of the population living more than one mile from a supermarket.The predictor variables include poverty rate, median family income, urban classification, low vehicle access, and SNAP low-access share. 


#### Histogram: Low Food Access Share

```{r}
# Create a histogram to examine the distribution of low food access

ggplot(clean_data, aes(x = low_access_share)) +
  geom_histogram(
    fill = brewer.pal(5, "Pastel1")[2],
    color = "black",
    bins = 30
  ) +
  labs(
    title = "Distribution of Low Food Access Across Census Tracts",
    x = "Percentage of Population with Low Food Access",
    y = "Count",
    caption = "Low food access represents the percentage of residents living more than one mile from a supermarket."
  ) +
  theme_minimal()
```
#### Histogram: Poverty Rate

```{r}
# Create a histogram to examine poverty rates

ggplot(clean_data, aes(x = poverty_rate)) +
  geom_histogram(
    fill = brewer.pal(5, "Pastel1")[3],
    color = "black",
    bins = 30
  ) +
  labs(
    title = "Distribution of Poverty Rate Across Census Tracts",
    x = "Poverty Rate (%)",
    y = "Count",
    caption = "Poverty rate represents the percentage of individuals living below the federal poverty line."
  ) +
  theme_minimal()
```

#### Histogram: Median Family Income

```{r}
# Create a histogram to examine median family income

ggplot(clean_data, aes(x = median_income)) +
  geom_histogram(
    fill = brewer.pal(5, "Pastel1")[4],
    color = "black",
    bins = 30
  ) +
  labs(
    title = "Distribution of Median Family Income",
    x = "Median Income (USD)",
    y = "Count",
    caption = "Median income represents the midpoint of household income distribution within each census tract."
  ) +
  theme_minimal()
```
#### Scatterplot: Poverty Rate vs. Low Food Access

```{r}
# Create a scatterplot to examine the relationship between poverty and low food access

ggplot(clean_data,
       aes(x = poverty_rate,
           y = low_access_share)) +
  geom_point(
    alpha = 0.4,
    color = brewer.pal(5, "Pastel1")[1]
  ) +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Poverty Rate vs Low Food Access",
    x = "Poverty Rate (%)",
    y = "Low Food Access (%)",
    caption = "This plot shows the relationship between poverty and the percentage of residents with limited access to supermarkets."
  ) +
  theme_minimal()
```

Scatterplot: Median Income vs Low Food Access

```{r}
# Create a scatterplot to examine the relationship between income and low food access

ggplot(clean_data,
       aes(x = median_income,
           y = low_access_share)) +
  geom_point(
    alpha = 0.4,
    color = brewer.pal(5, "Pastel1")[5]
  ) +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "Median Income vs Low Food Access",
    x = "Median Income (USD)",
    y = "Low Food Access (%)",
    caption = "This plot shows how income levels relate to access to supermarkets across census tracts."
  ) +
  theme_minimal()
```
#### Scatterplot: SNAP Low Access Share vs Lood Food Access

```{r}
# Create a scatterplot to examine SNAP low access share and low food access

ggplot(clean_data,
       aes(x = snap_low_access_share,
           y = low_access_share)) +
  geom_point(
    alpha = 0.4,
    color = brewer.pal(5, "Pastel1")[1]
  ) +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title = "SNAP Low Access Share vs Low Food Access",
    x = "SNAP Low Access Share (%)",
    y = "Low Food Access (%)",
    caption = "This plot shows the relationship between SNAP households with low access and overall low food access."
  ) +
  theme_minimal()
```
#### Boxplot: Low Food Access by Urban vs Rural Classification

```{r}
# Create a boxplot comparing low food access by urban classification

ggplot(clean_data,
       aes(x = urban,
           y = low_access_share,
           fill = urban)) +
  geom_boxplot() +
  scale_x_discrete(
    labels = c(
      "0" = "Rural",
      "1" = "Urban"
    )
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Low Food Access by Urban vs Rural Areas",
    x = "Urban Classification",
    y = "Low Food Access (%)",
    caption = "This plot compares the distribution of low food access between urban and rural census tracts."
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
#### Boxplot: Low Food Access by Vehicle Access

```{r}
# Create a boxplot comparing low food access by vehicle access

ggplot(clean_data,
       aes(x = low_vehicle_access,
           y = low_access_share,
           fill = low_vehicle_access)) +
  geom_boxplot() +
  scale_x_discrete(
    labels = c(
      "0" = "No Low Vehicle Access",
      "1" = "Low Vehicle Access"
    )
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Low Food Access by Vehicle Access",
    x = "Vehicle Access Status",
    y = "Low Food Access (%)",
    caption = "This plot compares low food access between tracts with and without low vehicle access."
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
`






